---
title: "CvH 03 — Differential Expression"
subtitle: "Design, contrasts, model fit, and the Π-score, on the non-imputed matrix"
author: "A_CvH_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here)
knitr::opts_knit$set(root.dir = here::here())
```

This document reproduces the primary DEP stage and shows the result of each step.
It runs the same four-step proteoDA workflow as
`03_DEP/a_non_imputed/a_script/01_run_dep.R` — describe the design, define the
comparisons, fit the model, pull the results — live and in memory. The headline
counts are computed from the live fit, then checked against the committed output.

# Start from the non-imputed matrix {#sec-load}

The test runs on the matrix that still has its missing values; `limma` handles
NAs per protein, which is the reason stage 02 kept the non-imputed object as the
primary hand-off:

```{r}
#| label: load
pacman::p_load(proteoDA)
dal <- readRDS(here("02_Normalization", "c_data", "DAList_normalized.rds"))
dim(dal$data)
table(dal$metadata$group_time)            # three groups, 39 samples
```

# Step 1 — describe the design {#sec-design}

`add_design()` turns a formula into a model matrix. `~ 0 + group_time` gives each
group its own column (its own mean), which makes the contrasts read cleanly.
`(1 | Subject_ID)` encodes that each subject appears at two timepoints:

```{r}
#| label: add-design
dal <- add_design(dal, "~ 0 + group_time + (1 | Subject_ID)")
colnames(dal$design)
head(dal$design)
```

The `(1 | Subject_ID)` term is not a column in the design matrix; proteoDA holds
it and passes it to `limma::duplicateCorrelation` at the fitting step.

# Step 2 — define the comparisons {#sec-contrasts}

Each contrast is a name and a difference of group means:

```{r}
#| label: add-contrasts
CONTRASTS <- c("CRvH_Baseline = CR_pre - H_pre",     # D: disease deviation from healthy
               "CR_Training   = CR_post - CR_pre",   # T: effect of training
               "Resid         = CR_post - H_pre")    # R: what remains after training (R = D + T)
dal <- add_contrasts(dal, contrasts_vector = CONTRASTS)
dal$contrasts
```

Read each column downward to see the group means being differenced — for
`CRvH_Baseline`, `+1` on `CR_pre` and `−1` on `H_pre`.

## What the three contrasts mean — and what `Resid` is *not* {#sec-contrast-meaning}

For one protein at three points — **healthy (H)**, **cancer-recovery before
training (CR_pre)**, **cancer-recovery after training (CR_post)** — the contrasts
are the three differences:

- **`CRvH_Baseline` = CR_pre − H** — the **disease** effect (call it **D**): how
  far the cancer-recovery state pushed the protein from healthy, before training.
- **`CR_Training` = CR_post − CR_pre** — the **training** effect (**T**): how far
  training moved it.
- **`Resid` = CR_post − H** — the **residual**: how far it is still from healthy
  after training. `Resid = D + T`.

`Resid` is **not** the reversal analysis. It is the leftover gap — a plain
difference. The reversal analysis is a separate downstream step
(`04_Figures/Reversal/`) that asks whether training moved the protein back *toward*
healthy, via a per-protein reversal fraction:

$$\varphi = -\,T / D$$

| φ | What happened |
|---|---|
| **0** | training did nothing (T = 0) — stuck at the diseased level |
| **0 to 1** | partial reversal — training undid some of the disease change |
| **1** | training exactly cancelled the disease (T = −D) — back to healthy |
| **> 1** | overshoot — past healthy to the other side |
| **< 0** | training made it worse — further from healthy |

The two views are tied by **`Resid = D · (1 − φ)`**: φ = 1 gives zero residual
gap, φ = 0 leaves the whole gap. `Resid` is the leftover gap (a difference); φ is
the fraction of the way back (a ratio).

# Step 3 — fit the model {#sec-fit}

One call estimates a single within-subject correlation, fits the model with it,
and applies robust empirical-Bayes moderation:

```{r}
#| label: fit
dal <- fit_limma_model(dal)
dal$eBayes_fit$correlation     # consensus within-subject correlation
```

That one consensus correlation lets a single model respect the paired design
without spending a degree of freedom per subject.

# Step 4 — pull the results {#sec-extract}

`extract_DA_results()` builds a tidy table per contrast into `$results`:

```{r}
#| label: extract
dal <- extract_DA_results(dal, pval_thresh = 0.10, lfc_thresh = 0, adj_method = "BH")
names(dal$results)
head(dal$results$CRvH_Baseline[, c("logFC", "P.Value", "adj.P.Val")])
```

# The Π-score {#sec-pi}

A small p-value on a barely-moving protein is not interesting. The Π-score
(Xiao 2014, $\Pi = \text{P.Value}^{|\text{logFC}|}$) folds effect size into
significance, so it is only small when a protein is both significant and moves a
lot. Each protein is tagged `+1` (up), `−1` (down), or `0` at Π < 0.05:

```{r}
#| label: pi
pacman::p_load(purrr, tibble, dplyr)
res <- imap(dal$results, \(r, cn) as_tibble(r, rownames = "uniprot_id") |>
  mutate(contrast = cn,
         pi_score = P.Value ^ abs(logFC),
         sig_pi = case_when(pi_score < 0.05 & logFC > 0 ~ 1L,
                            pi_score < 0.05 & logFC < 0 ~ -1L, TRUE ~ 0L)))
res$CRvH_Baseline |> dplyr::count(sig_pi)   # disease axis: up (+1) / down (-1) / not-sig (0)
```

# Headline numbers {#sec-results}

Summary across the three contrasts, computed from the live fit:

```{r}
#| label: summary
bind_rows(res) |>
  group_by(contrast) |>
  summarise(n = n(),
            FDR_0.10 = sum(adj.P.Val < 0.10, na.rm = TRUE),
            Pi_0.05  = sum(sig_pi != 0, na.rm = TRUE), .groups = "drop")
```

The disease axis has hundreds of hits; `CR_Training` has **zero** at FDR —
training is real but spread thinly across many proteins, none individually
decisive. That is why the reversal analysis works at the *set* level rather than
chasing individual training hits (`CvH_pipeline.qmd §6`).

The production script additionally runs proteoDA's native writers (shown here)
and saves `results.xlsx` + `combined_results_pi.csv` to `c_data`, with volcano /
MD / p-value-histogram plots to `b_reports`:

```{r}
#| label: writers
#| eval: false
write_limma_tables(dal, output_dir = here("03_DEP", "a_non_imputed", "c_data"),
                   annot_cols = c("uniprot_id", "gene", "protein", "description"), overwrite = TRUE)
write_limma_plots(dal, grouping_column = "group_time",
                  output_dir = here("03_DEP", "a_non_imputed", "b_reports"), overwrite = TRUE)
```

The live summary matches the committed `combined_results_pi.csv`:

```{r}
#| label: verify
pacman::p_load(readr)
read_csv(here("03_DEP", "a_non_imputed", "c_data", "combined_results_pi.csv"),
         show_col_types = FALSE) |>
  group_by(contrast) |>
  summarise(n = n(), FDR_0.10 = sum(adj.P.Val < 0.10, na.rm = TRUE),
            Pi_0.05 = sum(sig_pi != 0, na.rm = TRUE), .groups = "drop")
```

# Did imputation matter? {#sec-concordance}

`03_DEP/b_imputed/` re-runs this model on the two imputed matrices and correlates
the fold changes against the non-imputed answer:

```{r}
#| label: concordance
read_csv(here("03_DEP", "b_imputed", "c_data", "logFC_vs_nonimputed.csv"),
         show_col_types = FALSE) |> as.data.frame()
```

The strong contrasts agree almost perfectly (ρ ≈ 0.9+); only the weak training
axis wobbles between methods — another reason to report the non-imputed result.

# Recap {#sec-recap}

1. `add_design()` — a column per group, plus the subject block.
2. `add_contrasts()` — name-equals-difference comparisons.
3. `fit_limma_model()` — one consensus correlation + robust eBayes.
4. `extract_DA_results()` — tidy `$results`, one table per contrast.
5. The Π-score for effect-size-aware calls, then proteoDA's native writers.

The test runs on the non-imputed matrix; the imputed runs are a concordance
check. From here, `04_Figures/Reversal/` turns these three contrasts into the
reversal analysis.
