---
title: "CvH 01 — Filtering"
subtitle: "The DAList object; contaminant, missingness, and outlier removal"
author: "A_CvH_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here)
knitr::opts_knit$set(root.dir = here::here())
```

This document reproduces stage 01 of the CvH pipeline and shows the result of
each step. It runs the same logic as `01_Filtering/a_script/01_run_filtering.R`,
but entirely in memory: it computes the filtered object live and writes no files.
The final section confirms the live result matches the committed
`DAList_filtered.rds`.

# Load the raw matrix and sample sheet {#sec-load}

Stage 01 begins with a raw DIA-MS spreadsheet (`00_input/CvH_raw.xlsx`). The
annotation columns (protein identity) are separated from the sample-intensity
columns. The sample sheet `00_input/CvH_meta.csv` supplies one row per sample;
the pooled three-group factor `group_time` is derived from `Group_Time`, pooling
the creatine and placebo arms because the supplement is not the question here.
The matrix is aligned to the sheet by `Col_ID`.

```{r}
#| label: load
pacman::p_load(proteoDA, readxl, readr, dplyr, tidyr, stringr)

raw <- read_excel(here("00_input", "CvH_raw.xlsx"))
annot_cols <- c("uniprot_id", "protein", "gene", "description", "n_seq")
annotation <- raw[, annot_cols]
intensity  <- raw[, setdiff(names(raw), annot_cols)]

metadata <- as.data.frame(read_csv(here("00_input", "CvH_meta.csv"), show_col_types = FALSE))
metadata$group_time <- case_when(
  metadata$Group_Time %in% c("CRE_T1", "PLA_T1") ~ "CR_pre",
  metadata$Group_Time %in% c("CRE_T2", "PLA_T2") ~ "CR_post",
  metadata$Group_Time == "H_T1"                  ~ "H_pre")
stopifnot(!anyNA(metadata$group_time))
rownames(metadata) <- metadata$Col_ID
intensity <- intensity[, metadata$Col_ID]    # errors if matrix/metadata IDs disagree
n_raw <- nrow(annotation)

c(proteins = n_raw, samples = ncol(intensity))
table(metadata$group_time)                   # healthy baseline, CR pre-, CR post-training
```

# Step 1: remove blood contaminants {#sec-contaminants}

Muscle biopsies carry blood, and plasma proteins (albumin, hemoglobins,
immunoglobulins) are abundant enough to drown out the muscle signal. A protein is
flagged for removal if it looks like blood (HPA "Secreted to blood", an
immunoglobulin, or a strong erythrocyte signal) **and** does not look like muscle
(rescued by a strong skeletal-myonuclei signal). The classifier is built from the
Human Protein Atlas annotations:

```{r}
#| label: classifier
bl <- read_tsv(here("00_input", "HPA_annotations.tsv"), show_col_types = FALSE) |>
  transmute(
    gene          = Gene,
    uniprot       = Uniprot,
    protein_class = `Protein class`,
    secretome     = `Secretome location`,
    blood_conc    = suppressWarnings(as.numeric(`Blood concentration - Conc. blood MS [pg/L]`)),
    ery           = suppressWarnings(as.numeric(`Single Cell Type RNA - Erythrocytes [nCPM]`)),
    myo           = suppressWarnings(as.numeric(`Single Cell Type RNA - Myonuclei [nCPM]`))) |>
  mutate(
    secreted_blood = secretome == "Secreted to blood" & !is.na(secretome),
    is_ig          = str_detect(coalesce(protein_class, ""), "Immunoglobulin genes"),
    is_erythrocyte = !is.na(ery) & ery >= 5000,   # hemoglobin-class red-cell protein
    muscle_keep    = !is.na(myo) & myo >= 50,      # myonuclei signal rescues as real muscle
    reason = case_when(
      muscle_keep & (secreted_blood | is_ig | is_erythrocyte) ~ "keep: muscle-expressed (rescued)",
      is_erythrocyte                                          ~ "remove: erythrocyte (hemoglobin)",
      is_ig                                                   ~ "remove: immunoglobulin",
      secreted_blood                                          ~ "remove: secreted-to-blood (plasma)",
      TRUE                                                    ~ "keep: not blood-associated"),
    verdict = if_else(str_starts(reason, "remove"), "remove", "keep"))

table(bl$verdict)                              # how the HPA genes split
```

Removal is plain subsetting, applied to `annotation` and `intensity` together so
they stay aligned. The cascade keeps only proteins present in HPA, then drops the
blood-flagged ones; each step is logged:

```{r}
#| label: apply-contaminants
hpa_genes    <- unique(bl$gene)
remove_genes <- bl$gene[bl$verdict == "remove"]

flog <- tibble(step = "Raw input", n_after = nrow(annotation), n_removed = NA_integer_)
keep <- annotation$gene %in% hpa_genes
flog <- bind_rows(flog, tibble(step = "HPA presence", n_after = sum(keep), n_removed = sum(!keep)))
annotation <- annotation[keep, ]; intensity <- intensity[keep, ]
keep <- !(annotation$gene %in% remove_genes)
flog <- bind_rows(flog, tibble(step = "Blood contaminant removal", n_after = sum(keep), n_removed = sum(!keep)))
annotation <- annotation[keep, ]; intensity <- intensity[keep, ]

flog
```

The removed and rescued proteins, with the HPA evidence behind each call:

```{r}
#| label: removed-rescued
bl |> filter(verdict == "remove") |>
  arrange(reason, desc(blood_conc)) |>
  select(gene, reason, blood_conc, ery, myo) |> head(8)

bl |> filter(str_detect(reason, "rescued")) |>
  arrange(desc(myo)) |>
  select(gene, reason, blood_conc, myo) |> head(6)   # e.g. A2M: looks blood-like, kept as muscle
```

# Build the `DAList` {#sec-build}

proteoDA operates on a `DAList` — a named list holding three name-aligned tables:
`data` (intensity matrix), `annotation` (one row per protein), and `metadata`
(one row per sample). The row names of `data` equal the row names of
`annotation`, and the column names of `data` equal the row names of `metadata`.
`zero_to_missing()` turns DIA zeros into `NA`, since a reported zero means "not
detected" rather than a measured zero:

```{r}
#| label: build-dalist
int_mat  <- as.data.frame(data.matrix(intensity)); rownames(int_mat) <- annotation$uniprot_id
annot_df <- as.data.frame(annotation);             rownames(annot_df) <- annotation$uniprot_id
meta_df  <- as.data.frame(metadata);               rownames(meta_df)  <- metadata$Col_ID

dal <- zero_to_missing(DAList(data = int_mat, annotation = annot_df, metadata = meta_df))

identical(rownames(dal$data), rownames(dal$annotation))   # proteins aligned
identical(colnames(dal$data), rownames(dal$metadata))     # samples aligned
round(100 * mean(is.na(dal$data)), 1)                      # percent missing after zero -> NA
```

# Step 2: drop barely-detected proteins {#sec-missingness}

`filter_proteins_by_group()` is the native proteoDA filter. With `min_reps = 5,
min_groups = 1` it keeps a protein detected in at least 5 samples in **at least
one** group — a protein switched on in cancer but off in healthy is kept, since
detection is not required everywhere:

```{r}
#| label: missingness
n0 <- nrow(dal$data)
dal <- filter_proteins_by_group(dal, min_reps = 5, min_groups = 1, grouping_column = "group_time")
flog <- bind_rows(flog, tibble(step = "Missingness (>=5 in >=1 group)",
                               n_after = nrow(dal$data), n_removed = n0 - nrow(dal$data)))
flog <- flog |> mutate(pct_of_raw = round(n_after / n_raw * 100, 1))
flog                                          # full cascade, raw -> final, with % of raw retained
```

# Step 3: detect and remove outlier samples {#sec-outliers}

Whether a whole sample is untrustworthy is decided by a consensus of four
independent heuristics — high missingness (with a paired pre→post arm), off
intensity level (Hampel 3×MAD), multivariate position in PCA space, and low
correlation with the other samples. A sample is dropped only if at least **three
of four** agree:

```{r}
#| label: outliers
m <- dal$metadata; pct_missing <- colMeans(is.na(dal$data)) * 100
paired <- m |> filter(Group != "PPS") |> count(Subject_ID) |> filter(n == 2) |> pull(Subject_ID)
delta <- setNames(rep(NA_real_, nrow(m)), m$Col_ID)
for (s in paired) {
  r <- m[m$Subject_ID == s, ]; t1 <- r$Col_ID[r$Timepoint == "T1"]; t2 <- r$Col_ID[r$Timepoint == "T2"]
  if (length(t1) == 1 && length(t2) == 1) delta[c(t1, t2)] <- abs(pct_missing[t2] - pct_missing[t1])
}
miss_thr <- quantile(pct_missing, 0.75) + 1.5 * IQR(pct_missing)                  # Tukey upper fence
dv <- delta[!is.na(delta)]; delta_thr <- if (length(dv) > 2) quantile(dv, 0.75) + 1.5 * IQR(dv) else Inf
miss_flag <- pct_missing > miss_thr | (!is.na(delta) & delta > delta_thr)
sm <- apply(log2(dal$data), 2, median, na.rm = TRUE)
mad_flag <- abs(sm - median(sm)) > 3 * mad(sm)                                    # Hampel 3xMAD
pc3 <- prcomp(t(log2(dal$data[rowSums(is.na(dal$data)) == 0, ])), center = TRUE, scale. = TRUE)$x[, 1:3]
pca_flag <- mahalanobis(pc3, colMeans(pc3), cov(pc3)) > qchisq(0.99, df = 3)      # chi-sq tail on PC1-3
cm <- cor(log2(dal$data), use = "pairwise.complete.obs")
mc <- apply(cm, 2, function(x) median(x[x < 1], na.rm = TRUE))                    # x<1 drops self-correlation
cor_flag <- mc < median(mc) - 3 * mad(mc)
outlier_diag <- tibble(Col_ID = colnames(dal$data), miss_flag = miss_flag[Col_ID], mad_flag = mad_flag[Col_ID],
                       pca_flag = pca_flag[Col_ID], cor_flag = cor_flag[Col_ID]) |>
  mutate(n_flags = miss_flag + mad_flag + pca_flag + cor_flag, consensus_outlier = n_flags >= 3)
outlier_ids <- outlier_diag$Col_ID[outlier_diag$consensus_outlier]

head(outlier_diag[order(-outlier_diag$n_flags), ], 6)
outlier_ids                                   # CR9_T1 and CR10_T1 trip >= 3 flags
```

```{r}
#| label: apply-outliers
if (length(outlier_ids)) dal <- filter_samples(dal, !(Col_ID %in% outlier_ids))
dim(dal$data)
```

# Result and verification {#sec-recap}

proteoDA logs every operation in the `tags` slot. Stage 01 turns a 3,172 × 41
spreadsheet into a clean **2,555 × 39** `DAList`. The live result matches the
committed `DAList_filtered.rds` exactly:

```{r}
#| label: verify
names(dal$tags)
dim(dal$data)

committed <- readRDS(here("01_Filtering", "c_data", "DAList_filtered.rds"))
c(same_dim       = identical(dim(dal$data), dim(committed$data)),
  same_proteins  = identical(sort(rownames(dal$data)), sort(rownames(committed$data))),
  same_samples   = identical(sort(colnames(dal$data)), sort(colnames(committed$data))))
```

The side-effecting script additionally writes `filtering_report.xlsx` (the
classifier verdicts, filter cascade, and outlier diagnostics shown above) and a
contaminant plot to `b_reports/`, then saves `DAList_filtered.rds` for stage 02.
Key points:

- A `DAList` is three name-aligned tables; keeping them in step is what proteoDA
  relies on. The native filters preserve the alignment; hand-subsetting must move
  `data` and `annotation` together.
- `zero_to_missing()` reflects that DIA zeros are missing values.
- `filter_proteins_by_group()` and `filter_samples()` are the native filters.

Stage 02 picks up this file and normalizes it.
