---
title: "CvH Proteomics — Imputation Pipeline"
subtitle: "MAR/MNAR Classification, Benchmark, and Quality Audit"
author: "DTL"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 7
execute:
  warning: false
  message: false
---

# 0 -- Setup {#sec-setup}

This document presents the results of the CvH imputation pipeline. It
does not execute the imputation analysis itself --- the companion script
`CvH_imputation_run.R` produces all data outputs and report PDFs. This
narrative document reads those outputs and contextualizes them within the
proteomics literature on missing data handling. All code chunks are
display-only: they read pre-computed CSV files and generate tables or
figures. No imputation computation is performed here.

```{r setup}

if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(readr, dplyr, tidyr, knitr, ggplot2, patchwork, here, scales)

# Paths to pre-computed outputs from CvH_imputation_run.R
data_dir   <- file.path(here::here(), "02_Imputation", "c_data")
report_dir <- file.path(here::here(), "02_Imputation", "b_reports")
norm_dir   <- file.path(here::here(), "01_normalization", "c_data")
meta_path  <- file.path(here::here(), "00_input", "CvH_meta.csv")

# Palettes consistent with the R script
pal_mar <- c(MAR = "#4393C3", MNAR = "#D6604D", Complete = "#999999")
pal_typ <- c(MNAR = "#D6604D", MAR = "#4393C3", Hybrid = "#5AAE61")
pal_donut <- c(Complete = "#999999", MAR = "#4393C3", MNAR = "#D6604D")
pal_bio <- c(CR_T1 = "#4393C3", CR_T2 = "#2166AC", PPS_T1 = "#B2182B")
```


# 1 -- Introduction: The Missing Data Problem in DIA-MS {#sec-intro}

Missing values are pervasive in label-free quantitative proteomics.
A typical DIA-MS intensity matrix retains 5--20% missing entries even
after quality filtering, and the manner in which those gaps are handled
directly affects fold changes, variance estimates, p-values, and false
discovery rates. Ignoring the problem --- by analyzing only complete
cases or by relying on a default imputation strategy without evaluation
--- introduces systematic bias whose direction and magnitude depend on
the underlying missingness mechanism (Lazar et al., 2016).

The statistical framework for missing data distinguishes three
mechanisms: values missing completely at random (MCAR), missing at
random (MAR), and missing not at random (MNAR). In proteomics, MNAR is
particularly important because it encompasses the dominant mode of data
loss in mass spectrometry: peptides whose abundance falls below the
detection limit produce no measurable signal, creating a systematic
association between low intensity and missingness (Webb-Robertson et
al., 2015).

DIA-MS acquires fragmentation spectra across the entire precursor mass
range in each cycle, substantially reducing the stochastic missingness
that plagues DDA. In DDA, the mass spectrometer selects a limited number
of precursors per scan cycle, generating MAR-like missingness independent
of peptide abundance. DIA eliminates this randomness by fragmenting all
precursors within predefined isolation windows. The practical consequence
is that missing values in DIA data are overwhelmingly MNAR --- they
arise because the peptide was genuinely below quantification threshold,
not because the instrument selected other precursors (Kong et al., 2022).

This distinction has direct implications for method selection. MAR-focused
methods (knn, bpca) exploit the correlation structure among observed
values but systematically overestimate truly low-abundance missing
entries because they predict from the center of the observed distribution
(Lazar et al., 2016). MNAR-focused methods (MinProb, QRILC) draw
replacement values from the lower tail, appropriately modeling
below-detection censoring but underestimating values that are genuinely
MAR (Webb-Robertson et al., 2015).

Because real datasets contain a mixture of both mechanisms, a single
method applied uniformly is unlikely to be optimal. The mixed-imputation
approach addresses this by first classifying each protein as MAR or MNAR,
then applying the appropriate method to each subset (Lazar et al., 2016).
Whether a mixed approach outperforms a well-chosen single method depends
on the proportion of MAR versus MNAR proteins and the accuracy of the
classification.

Given this complexity, benchmarking is essential. There is no universally
best imputation method; the optimal choice depends on the missingness
pattern, data structure, and downstream goals. A method that minimizes
reconstruction error on one dataset may fail on another with different
characteristics. Dataset-specific evaluation is the only reliable way to
make this determination (Jin et al., 2021).

Our pipeline proceeds in three stages: (1) classify each protein as MAR
or MNAR using msImpute `selectFeatures`; (2) benchmark 14 methods using
NRMSE on 10% masked values across 20 iterations; (3) apply the best
method and audit imputation quality. The distinction between benchmarking
and blind application is consequential --- many pipelines apply a default
without evaluating appropriateness, which is problematic for mixed-mechanism
datasets (Harris et al., 2023).

The imputation step sits between normalization (stage 01) and
differential expression (stage 04). Its outputs feed into a sensitivity
analysis (stage 03) that compares results across normalization and
imputation combinations, providing an additional check on whether the
imputation choice materially affects biological conclusions.

The CvH study compares cancer survivors (CR) who completed 12 weeks of
resistance training against healthy age-matched controls (PPS), with CR
participants randomized to creatine or placebo. Vastus lateralis biopsies
were collected at baseline (T1) and post-training (T2) for CR, and at a
single timepoint for controls. Condition-specific proteins may be
systematically missing in one group, reinforcing the expectation that
MNAR dominates the missingness pattern.


# 2 -- MAR/MNAR Classification {#sec-classification}

Before selecting an imputation method, it is important to characterize
the missingness mechanism. The distinction between MAR and MNAR
determines which class of method is appropriate, and applying the wrong
class introduces systematic bias whose direction is predictable from the
mismatch (Lazar et al., 2016).

The msImpute package provides `selectFeatures`, which fits an
evidence-based model (EBM) to distinguish proteins whose missingness
pattern is consistent with random dropout (MAR) from those consistent
with abundance-dependent censoring (MNAR). The function evaluates
whether missing value patterns can be explained by observed intensity
structure or require a left-censoring model (Wei et al., 2018).

When `selectFeatures` fails --- due to insufficient patterns for model
fitting --- the pipeline falls back to an intensity heuristic: proteins
with >30% missingness and below-median intensity, or >50% missingness
regardless of intensity, are classified as MNAR. This heuristic captures
the well-documented inverse relationship between abundance and
missingness (Webb-Robertson et al., 2015). Proteins with no missing
values are classified as Complete.

The classification serves two purposes. First, it informs the mixed
imputation methods in the benchmark, which apply different algorithms to
the MAR and MNAR subsets. Second, it provides interpretive context: if
the dataset is predominantly MNAR, we expect MNAR-focused or hybrid
methods to outperform pure MAR methods (Lazar et al., 2016).

```{r mar-mnar-table}

miss_class <- read_csv(file.path(data_dir, "mar_mnar_classification.csv"),
                       show_col_types = FALSE)

class_summary <- miss_class %>%
  count(class) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  arrange(desc(n))

kable(class_summary,
      col.names = c("Class", "N Proteins", "% of Total"),
      caption = "MAR/MNAR classification summary")
```

The donut chart below provides a visual summary of the classification
breakdown. Each slice represents the proportion of proteins assigned to
one of three categories: Complete (no missing values), MAR (missingness
unrelated to abundance), or MNAR (missingness driven by low abundance).
The relative sizes immediately convey whether the dataset is dominated by
one mechanism, which in turn guides expectations for benchmark results.

```{r donut-chart}
#| fig-cap: "Proportion of proteins classified as Complete, MAR, or MNAR"

donut_df <- class_summary %>%
  mutate(
    fraction = n / sum(n),
    ymax     = cumsum(fraction),
    ymin     = lag(ymax, default = 0),
    label_y  = (ymin + ymax) / 2,
    label    = paste0(class, "\n", n, " (", pct, "%)")
  )

ggplot(donut_df, aes(ymax = ymax, ymin = ymin, xmax = 4, xmin = 2.5, fill = class)) +
  geom_rect(alpha = 0.85, color = "white", linewidth = 0.8) +
  geom_text(aes(x = 3.25, y = label_y, label = label), size = 3.5, fontface = "bold") +
  scale_fill_manual(values = pal_donut) +
  coord_polar(theta = "y") +
  xlim(c(1, 4.5)) +
  labs(title = "MAR / MNAR Classification of Proteins",
       fill = "Class") +
  theme_void(base_size = 12) +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

The plot below shows the relationship between a protein's average
observed intensity and the percentage of samples in which it is missing.
Proteins that are abundant tend to be detected reliably across all
samples, while low-abundance proteins are more frequently missing. This
pattern is the hallmark of MNAR missingness: values are absent precisely
because the protein's signal falls below the instrument's detection
limit, not by random chance. The loess smooth line highlights this trend.

```{r missingness-vs-intensity}
#| fig-cap: "Protein missingness percentage versus mean observed intensity, colored by MAR/MNAR class"

miss_scatter <- miss_class %>%
  filter(!is.na(mean_int), !is.na(pct_miss))

ggplot(miss_scatter, aes(x = mean_int, y = pct_miss, color = class)) +
  geom_point(alpha = 0.55, size = 1.8) +
  geom_smooth(method = "loess", se = TRUE, color = "grey30",
              linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = pal_mar) +
  labs(x = "Mean observed intensity (log2)",
       y = "% missing values",
       title = "Missingness increases as protein abundance decreases",
       color = "Class") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The bar chart below shows how many values are observed versus missing in
each individual sample. This sample-level view complements the
protein-level classification above. Samples with unusually high
missingness may indicate technical issues (poor digestion, low loading),
while consistent missingness across samples supports a protein-driven
(MNAR) explanation. Bars are ordered by biological group so that
patterns within and between conditions are visible.

```{r sample-missingness-bar}
#| fig-cap: "Per-sample observed and missing value counts, ordered by biological group"

norm_mat <- read_csv(file.path(norm_dir, "01_normalized.csv"),
                     show_col_types = FALSE)
meta     <- read_csv(meta_path, show_col_types = FALSE)

# Identify sample columns (exclude annotation columns)
sample_cols <- setdiff(names(norm_mat),
                       c("uniprot_id", "protein", "gene", "description", "n_seq"))

sample_miss <- tibble(
  Col_ID   = sample_cols,
  n_miss   = colSums(is.na(norm_mat[, sample_cols])),
  n_obs    = nrow(norm_mat) - colSums(is.na(norm_mat[, sample_cols]))
) %>%
  left_join(meta %>% select(Col_ID, Group_Time), by = "Col_ID") %>%
  mutate(BioGroup = case_when(
    Group_Time %in% c("CRE_T1", "PLA_T1") ~ "CR_T1",
    Group_Time %in% c("CRE_T2", "PLA_T2") ~ "CR_T2",
    Group_Time == "H_T1"                   ~ "PPS_T1",
    TRUE                                   ~ Group_Time
  )) %>%
  arrange(BioGroup, Col_ID)

sample_miss_long <- sample_miss %>%
  pivot_longer(cols = c(n_obs, n_miss),
               names_to = "status", values_to = "count") %>%
  mutate(status = factor(status, levels = c("n_obs", "n_miss"),
                         labels = c("Observed", "Missing")),
         Col_ID = factor(Col_ID, levels = sample_miss$Col_ID))

ggplot(sample_miss_long, aes(x = Col_ID, y = count, fill = status)) +
  geom_col(width = 0.75) +
  scale_fill_manual(values = c(Observed = "#4393C3", Missing = "#D6604D")) +
  labs(x = NULL, y = "Number of proteins",
       title = "Per-sample observed vs. missing protein counts",
       fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7),
        legend.position = "bottom")
```

For DIA-MS data, we expect the MNAR fraction to dominate among proteins
with missing values, since the primary source of data loss is peptide
abundance falling below the detection limit rather than stochastic
precursor selection. If the MNAR fraction substantially exceeds MAR, this
confirms alignment with the expected left-censoring model. The relative
proportions also inform benchmark expectations: an MNAR-dominated dataset
should favor methods drawing from the lower intensity tail.

The intensity distributions of MAR versus MNAR proteins should show clear
separation, with MNAR proteins having lower mean observed intensity. This
validates the classification --- identical distributions would render the
distinction meaningless. The scatter plot above confirms this separation
visually.

Understanding the classification contextualizes the downstream quality
audit. MNAR-classified proteins will receive imputed values from the left
tail of their distribution, shifting their mean downward. This shift
should be small (Cohen's d < 0.2 for most proteins) if the imputation is
well-calibrated, but proteins with high missingness will inevitably show
larger shifts because more of their values are replaced (Wei et al.,
2018).


# 3 -- Benchmark Design and Rationale {#sec-benchmark-design}

Selecting an imputation method without benchmarking is equivalent to
choosing a statistical model without checking assumptions. Our benchmark
evaluates 14 candidate methods to identify which best reconstructs
artificially masked observed values in this specific dataset.

The 14 methods span three categories. Three MNAR-focused methods draw
from left-censored distributions: MinProb samples from a Gaussian
centered at the lowest quantile, MinDet uses the minimum observed value,
and QRILC fits a truncated Gaussian per protein. These are appropriate
when missing values represent below-detection measurements (Lazar et al.,
2016).

Four MAR-focused methods exploit correlation structure: knn imputes from
weighted averages of similar proteins, bpca uses probabilistic PCA, MLE
fits a multivariate Gaussian, and RF trains decision tree ensembles.
These assume missingness is unrelated to abundance (Webb-Robertson et
al., 2015).

Seven mixed methods combine one MAR and one MNAR method, applying each
to its classified protein subset: knn+MinProb, knn+MinDet, knn+QRILC,
bpca+MinProb, bpca+QRILC, RF+MinProb, and RF+QRILC. This tests whether
mechanism-aware imputation outperforms uniform application (Lazar et al.,
2016).

The benchmark masks 10% of observed values, imputes the full matrix, and
compares imputed values at masked positions to the known truth. This
masking rate balances stable error estimates against preserving enough
observed values for imputation algorithms to function (Jin et al., 2021).

Reconstruction accuracy is measured by NRMSE (RMSE / SD of true values),
a scale-invariant metric that directly quantifies reconstruction
accuracy. Values near zero indicate near-perfect recovery; values near
1.0 indicate error comparable to natural variability (Jin et al., 2021).

The masking procedure is repeated 20 times with different random masks
(seed = 42) to obtain stable distributions. During masking, proteins
gaining additional missingness are temporarily reclassified to prevent
MNAR methods from being evaluated on artificially masked high-abundance
values.

The design intentionally includes methods expected to perform poorly
under certain conditions because negative results are informative. If a
pure MNAR method outperforms all others, it provides strong evidence of
predominantly left-censored missingness. If a mixed method wins, the
classification captures meaningful mechanistic heterogeneity (Harris et
al., 2023).

Only the best-performing method (by mean NRMSE) is applied to the full
dataset. The sensitivity analysis in stage 03 tests whether alternative
strategies would change biological conclusions. No single method
consistently outperforms all others across datasets (Jin et al., 2021;
Harris et al., 2023), which is precisely why dataset-specific
benchmarking is essential.


# 4 -- Benchmark Results and Interpretation {#sec-benchmark-results}

The benchmark summary from `CvH_imputation_run.R` contains mean,
standard deviation, and median NRMSE for each of the 14 methods across
20 masking iterations.

```{r benchmark-table}

bench <- read_csv(file.path(data_dir, "benchmark_summary.csv"),
                  show_col_types = FALSE)

bench <- bench %>%
  mutate(type = case_when(
    method %in% c("MinProb", "MinDet", "QRILC") ~ "MNAR",
    method %in% c("knn", "bpca", "MLE", "RF")   ~ "MAR",
    TRUE                                         ~ "Hybrid"
  )) %>%
  arrange(mean)

kable(bench,
      digits = 4,
      col.names = c("Method", "Mean NRMSE", "SD", "Median NRMSE", "Type"),
      caption = "Benchmark results: 14 imputation methods ranked by mean NRMSE (20 iterations, 10% masking)")
```

```{r benchmark-barplot}
#| fig-cap: "Mean NRMSE across 14 imputation methods with standard deviation error bars"

bench_plot <- bench %>%
  mutate(method = factor(method, levels = method))

ggplot(bench_plot, aes(x = method, y = mean, fill = type)) +
  geom_col(alpha = 0.8, width = 0.7) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.25, linewidth = 0.5) +
  scale_fill_manual(values = c(MNAR = "#D6604D", MAR = "#4393C3", Hybrid = "#5AAE61")) +
  labs(x = NULL, y = "Mean NRMSE",
       title = "Imputation benchmark: Mean NRMSE by method",
       fill = "Method type") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

The ranking reveals whether single-mechanism or mixed-mechanism methods
are better suited to this dataset. If a pure MNAR method ranks first, it
suggests the dominant missingness is abundance-dependent left censoring
and that correlation structure provides insufficient additional
information. If a mixed method wins, the MAR/MNAR classification captures
genuine heterogeneity that improves reconstruction (Lazar et al., 2016).

The standard deviation across iterations reflects method stability. A
method with low mean but high SD may be sensitive to the specific masking
pattern. The ideal method combines low mean NRMSE with low SD, indicating
both accurate and consistent reconstruction.

Methods with very high NRMSE (above 1.0) are effectively worse than
random noise and serve as negative controls. The spread between the best
and worst methods quantifies the stakes of imputation method selection:
a large spread means a default choice without benchmarking risks
substantial bias.

The benchmark bar plot above provides detail on within-method variability.
Regardless of which method wins, the benchmark provides the empirical
justification for the imputation choice applied in the R script.


# 5 -- Imputation Quality Audit {#sec-audit}

Selecting the best method by benchmark NRMSE is necessary but not
sufficient. The benchmark evaluates reconstruction of artificially masked
observed values, which are not truly missing. The actual missing values
are predominantly low-abundance proteins below the detection limit,
whereas masked values are drawn from the full observed distribution. The
quality audit examines whether imputation produces a biologically
plausible result on the real missing values.

The audit evaluates three dimensions: (1) per-protein effect sizes
(Cohen's d) quantifying how much imputation shifts each protein's mean;
(2) the relationship between effect size and missingness percentage; and
(3) PCA and sample-level diagnostics assessing preservation of biological
group structure.

```{r audit-summary}

audit <- read_csv(file.path(data_dir, "imputation_effect_audit.csv"),
                  show_col_types = FALSE)

audit_stats <- audit %>%
  summarise(
    n_proteins   = n(),
    median_d     = round(median(effect_d, na.rm = TRUE), 4),
    mean_d       = round(mean(effect_d, na.rm = TRUE), 4),
    sd_d         = round(sd(effect_d, na.rm = TRUE), 4),
    q25_d        = round(quantile(effect_d, 0.25, na.rm = TRUE), 4),
    q75_d        = round(quantile(effect_d, 0.75, na.rm = TRUE), 4),
    pct_small    = round(mean(abs(effect_d) < 0.2, na.rm = TRUE) * 100, 1),
    pct_medium   = round(mean(abs(effect_d) >= 0.2 & abs(effect_d) < 0.5, na.rm = TRUE) * 100, 1),
    pct_large    = round(mean(abs(effect_d) >= 0.5, na.rm = TRUE) * 100, 1)
  )

kable(t(audit_stats),
      col.names = "Value",
      caption = "Summary of per-protein imputation effect sizes (Cohen's d)")
```

Proteins with higher missingness are expected to show larger effect sizes
because a greater fraction of their values are replaced. The reference
lines at d = +/-0.2 mark the conventional threshold for a small effect.

```{r audit-scatter}
#| fig-cap: "Imputation effect size (Cohen's d) versus percent missingness per protein"

ggplot(audit, aes(x = pct_miss, y = effect_d)) +
  geom_point(alpha = 0.5, size = 1.2, color = "#4393C3") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_hline(yintercept = c(-0.2, 0.2), linetype = "dotted", color = "grey50") +
  annotate("text", x = max(audit$pct_miss, na.rm = TRUE) * 0.85, y = 0.22,
           label = "|d| = 0.2 (small effect)", size = 3, color = "grey40") +
  labs(x = "% missing values", y = "Effect size (Cohen's d)",
       title = "Per-protein imputation effect: shift / pre-imputation SD") +
  theme_minimal(base_size = 12)
```

A distribution centered near zero with most values within the +/-0.2
band indicates the imputation preserves original intensity structure. A
negative skew is expected for MNAR-focused methods, which pull
post-imputation means downward.

```{r audit-histogram}
#| fig-cap: "Distribution of per-protein imputation effect sizes (Cohen's d)"

ggplot(audit, aes(x = effect_d)) +
  geom_histogram(bins = 50, fill = "#4393C3", alpha = 0.7, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-0.2, 0.2), linetype = "dotted", color = "grey50") +
  labs(x = "Cohen's d (imputation effect)", y = "Count",
       title = "Distribution of imputation effect sizes across proteins") +
  theme_minimal(base_size = 12)
```

Cohen's d below 0.2 is conventionally a small effect --- the shift is
less than one-fifth of the natural variability. If the vast majority of
proteins fall within this band, imputation is unlikely to distort
downstream differential expression. Proteins exceeding |d| = 0.5 warrant
scrutiny, as the imputation may have substantially altered their
intensity distribution.

For MNAR-focused methods, we expect a negative correlation between effect
size and missingness: more missing values means more imputed left-tail
draws, producing larger downward shifts. A linear relationship indicates
consistent behavior; nonlinearity could signal model misspecification.

The plot below shows PCA projections of the protein intensity matrix
before and after imputation, side by side. PCA reduces the
high-dimensional protein data to two summary axes that capture the
largest sources of variation. If imputation is well-behaved, the
biological group structure (CR at baseline, CR post-training, and healthy
controls) should be preserved: groups that were separated before
imputation should remain separated, and their relative positions should
not shift. Artifactual clustering introduced by imputation would appear
as groups merging or new separations appearing that were not present in
the pre-imputation projection.

```{r pca-pre-post}
#| fig-cap: "PCA of protein intensities before (left) and after (right) imputation, colored by biological group"
#| fig-width: 14
#| fig-height: 6

# Read pre- and post-imputation matrices
pre_mat  <- read_csv(file.path(norm_dir, "01_normalized.csv"),
                     show_col_types = FALSE)
post_mat <- read_csv(file.path(data_dir, "01_imputed.csv"),
                     show_col_types = FALSE)
meta_pca <- read_csv(meta_path, show_col_types = FALSE)

sample_cols_pca <- setdiff(names(pre_mat),
                           c("uniprot_id", "protein", "gene", "description", "n_seq"))

# Helper: compute PCA scores from a wide protein matrix
run_pca <- function(mat, samp_cols, fill_na = TRUE) {
  m <- as.matrix(mat[, samp_cols])
  if (fill_na) {
    # Median-fill per protein for pre-imputation
    row_med <- apply(m, 1, median, na.rm = TRUE)
    for (j in seq_len(ncol(m))) {
      nas <- is.na(m[, j])
      m[nas, j] <- row_med[nas]
    }
  }
  pca <- prcomp(t(m), center = TRUE, scale. = TRUE)
  scores <- as.data.frame(pca$x[, 1:2])
  scores$Col_ID <- samp_cols
  var_exp <- summary(pca)$importance[2, 1:2] * 100
  list(scores = scores, var_exp = var_exp)
}

pca_pre  <- run_pca(pre_mat,  sample_cols_pca, fill_na = TRUE)
pca_post <- run_pca(post_mat, sample_cols_pca, fill_na = FALSE)

# Attach BioGroup
attach_bio <- function(scores_df, meta_df) {
  scores_df %>%
    left_join(meta_df %>% select(Col_ID, Group_Time), by = "Col_ID") %>%
    mutate(BioGroup = case_when(
      Group_Time %in% c("CRE_T1", "PLA_T1") ~ "CR_T1",
      Group_Time %in% c("CRE_T2", "PLA_T2") ~ "CR_T2",
      Group_Time == "H_T1"                   ~ "PPS_T1",
      TRUE                                   ~ Group_Time
    ))
}

pre_scores  <- attach_bio(pca_pre$scores,  meta_pca)
post_scores <- attach_bio(pca_post$scores, meta_pca)

p_pre <- ggplot(pre_scores, aes(x = PC1, y = PC2, color = BioGroup)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.68, linetype = "dashed", linewidth = 0.6) +
  scale_color_manual(values = pal_bio) +
  labs(x = paste0("PC1 (", round(pca_pre$var_exp[1], 1), "%)"),
       y = paste0("PC2 (", round(pca_pre$var_exp[2], 1), "%)"),
       title = "Pre-imputation (median-filled)",
       color = "Group") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_post <- ggplot(post_scores, aes(x = PC1, y = PC2, color = BioGroup)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.68, linetype = "dashed", linewidth = 0.6) +
  scale_color_manual(values = pal_bio) +
  labs(x = paste0("PC1 (", round(pca_post$var_exp[1], 1), "%)"),
       y = paste0("PC2 (", round(pca_post$var_exp[2], 1), "%)"),
       title = "Post-imputation",
       color = "Group") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_pre + p_post + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

The plot below overlays the intensity distributions of observed values
(from the pre-imputation matrix) and imputed values (values that were
missing before imputation but present after). For MNAR-focused methods,
imputed values are expected to appear as a secondary peak or shoulder in
the lower tail of the intensity distribution, reflecting the assumption
that missing proteins were below the detection limit. If the imputed
values overlap substantially with the observed distribution rather than
sitting below it, the method may be overestimating missing values ---
a sign that an MAR method was applied to MNAR data.

```{r intensity-overlay}
#| fig-cap: "Intensity distribution: observed values versus imputed (filled-in) values"

pre_long  <- pre_mat  %>%
  pivot_longer(cols = all_of(sample_cols_pca),
               names_to = "sample", values_to = "intensity_pre")
post_long <- post_mat %>%
  pivot_longer(cols = all_of(sample_cols_pca),
               names_to = "sample", values_to = "intensity_post")

combined <- left_join(
  pre_long  %>% select(gene, sample, intensity_pre),
  post_long %>% select(gene, sample, intensity_post),
  by = c("gene", "sample")
)

observed_vals <- combined %>%
  filter(!is.na(intensity_pre)) %>%
  transmute(intensity = intensity_pre, source = "Observed")

imputed_vals <- combined %>%
  filter(is.na(intensity_pre) & !is.na(intensity_post)) %>%
  transmute(intensity = intensity_post, source = "Imputed")

overlay_df <- bind_rows(observed_vals, imputed_vals)

ggplot(overlay_df, aes(x = intensity, fill = source)) +
  geom_histogram(aes(y = after_stat(density)), bins = 80,
                 alpha = 0.6, position = "identity", color = "white", linewidth = 0.2) +
  scale_fill_manual(values = c(Observed = "#4393C3", Imputed = "#D6604D")) +
  labs(x = "Log2 intensity", y = "Density",
       title = "Observed vs. imputed intensity distributions",
       fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The plot below assesses whether imputation systematically shifts any
sample's overall intensity. For each sample, the mean log2 intensity is
computed before and after imputation and plotted against the identity
line (y = x). Points lying exactly on the identity line indicate zero
net shift; points below it indicate the imputation lowered the sample
mean (expected for MNAR methods that fill in low-tail values), and
points above indicate an upward shift. Small, consistent downward
deviations are expected and benign. Large or heterogeneous deviations
would suggest the imputation is differentially affecting certain samples,
which could introduce bias into group comparisons.

```{r sample-mean-shift}
#| fig-cap: "Pre- versus post-imputation sample means with identity line, colored by biological group"

pre_means <- colMeans(pre_mat[, sample_cols_pca], na.rm = TRUE)
post_means <- colMeans(post_mat[, sample_cols_pca], na.rm = TRUE)

shift_df <- tibble(
  Col_ID    = sample_cols_pca,
  pre_mean  = pre_means,
  post_mean = post_means
) %>%
  left_join(meta_pca %>% select(Col_ID, Group_Time), by = "Col_ID") %>%
  mutate(BioGroup = case_when(
    Group_Time %in% c("CRE_T1", "PLA_T1") ~ "CR_T1",
    Group_Time %in% c("CRE_T2", "PLA_T2") ~ "CR_T2",
    Group_Time == "H_T1"                   ~ "PPS_T1",
    TRUE                                   ~ Group_Time
  ))

ggplot(shift_df, aes(x = pre_mean, y = post_mean, color = BioGroup)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 3.5, alpha = 0.8) +
  scale_color_manual(values = pal_bio) +
  labs(x = "Pre-imputation sample mean (log2)",
       y = "Post-imputation sample mean (log2)",
       title = "Sample mean intensity: pre vs. post imputation",
       color = "Group") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

These diagnostics inform a binary decision: is the imputation trustworthy
for downstream differential expression, or should we default to a
non-imputed pathway? Trustworthy imputation shows: (1) most effect sizes
below |d| = 0.2, (2) a monotonic effect-size/missingness relationship,
(3) preserved PCA group structure, and (4) sample means near the identity
line.

The quality audit operates at the global level and cannot detect whether
specific biologically important proteins have been distorted. The
sensitivity analysis in stage 03 addresses this by comparing differential
expression results across multiple imputation strategies, identifying
proteins whose significance is sensitive to the imputation decision.


# 6 -- Conclusion {#sec-conclusion}

This document has presented the results of the CvH imputation pipeline,
which classifies missingness mechanisms, benchmarks 14 imputation
methods, applies the best-performing method, and audits imputation
quality through multiple complementary diagnostics.

The MAR/MNAR classification provides the mechanistic foundation for
method selection, confirming whether the missingness pattern aligns with
the expected left-censoring model for DIA-MS data. The benchmark
identifies the method that best reconstructs masked values, providing
empirical justification. The quality audit evaluates whether the
imputation preserves biological signal and quantifies per-protein effect
sizes.

The imputed matrix (`01_imputed.csv`) serves as the primary input for
differential expression in stage 04. The sensitivity analysis in stage 03
will compare results across a normalization-imputation grid to determine
whether the choice materially affects final biological conclusions. If
substantial discordance is found, the non-imputed pathway provides a
conservative fallback avoiding imputation-induced bias at the cost of
reduced power.


# References {#sec-refs}

1. Lazar C, Gatto L, Ferro M, et al. (2016). Accounting for the multiple
   natures of missing values in label-free quantitative proteomics data
   sets to compare imputation strategies. *Journal of Proteome Research*
   15(4):1116-1125.

2. Webb-Robertson BM, Wiber HK, Matzke MM, et al. (2015). Review,
   evaluation, and discussion of the challenges of missing value
   imputation for mass spectrometry-based label-free global proteomics.
   *Journal of Proteome Research* 14(3):920-930.

3. Kong W, Hui HWH, Peng H, et al. (2022). Dealing with missing values
   in proteomics data. *Proteomics* 23(23-24):e2200092.

4. Wei R, Wang J, Su M, et al. (2018). Missing value imputation approach
   for mass spectrometry-based metabolomics data. *Scientific Reports*
   8:663.

5. Jin L, Bi Y, Hu C, et al. (2021). A comparative study of evaluating
   missing value imputation methods in label-free proteomics. *Scientific
   Reports* 11:1760.

6. Valikangas T, Suomi T, Elo LL (2018). A systematic evaluation of
   normalization methods in quantitative label-free proteomics. *Briefings
   in Bioinformatics* 19(1):1-11.

7. Harris L, Fondrie WE, Oh S, Noble WS (2023). Evaluating proteomics
   imputation methods with improved criteria. *Journal of Proteome
   Research* 22(11):3427-3438.

8. O'Brien JJ, Bhatt DK, Engel JM, et al. (2018). Label-free
   quantitative proteomics analysis with implications for missingness and
   filtering strategies. *Annals of Applied Statistics* 12(4):2484-2510.


# Session Info {#sec-session}

```{r session-info}

sessionInfo()
```
