---
title: "CvH Proteomics — Differential Expression Analysis"
subtitle: "limma with arrayWeights, duplicateCorrelation, and Robust eBayes"
author: "DTL"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 7
execute:
  warning: false
  message: false
---

```{r setup}
#| label: setup

if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(readr, dplyr, tidyr, knitr, ggplot2, patchwork, ggrepel, here)

data_dir   <- file.path(here::here(), "04_DEP", "c_data")
report_dir <- file.path(here::here(), "04_DEP", "b_reports")
```


# 1 --- Introduction: Differential Expression in Proteomics

Differential expression analysis in proteomics (DEP) identifies proteins
whose abundance differs systematically between experimental conditions.
Unlike simple fold-change filtering, formal statistical testing accounts
for measurement variability across biological replicates and controls the
rate of false discoveries when thousands of proteins are tested
simultaneously. The goal is to move beyond descriptive abundance profiles
toward rigorous inference about which proteins respond to the conditions
under study.

In the context of label-free quantitative proteomics, each protein's
measured intensity is subject to multiple sources of variation: biological
variability between subjects, technical variability from sample
preparation and LC-MS acquisition, and the stochastic nature of
data-dependent or data-independent acquisition. A sound statistical
framework must model these sources explicitly rather than treating the
data as if each protein were measured in isolation.

The limma framework (Ritchie et al., 2015) has become the dominant
approach for small-sample proteomics experiments because it stabilizes
variance estimates through empirical Bayes shrinkage. Rather than
estimating a variance for each protein independently---an unreliable
procedure when only a handful of replicates are available---limma borrows
information across all proteins to produce a moderated variance estimate.
This shrinkage pulls extreme per-protein variances toward a common prior,
yielding more stable test statistics and substantially greater power to
detect true differences (Smyth, 2004).

An alternative framework, DEqMS (Zhu et al., 2020), extends limma by
modeling the relationship between variance and peptide count. In
proteomics, proteins identified by more peptides tend to have more
precise quantification, and DEqMS leverages this relationship to produce
count-specific variance priors. However, DEqMS requires a complete data
matrix---it cannot accommodate missing values because the
peptide-count-to-variance relationship breaks down when some samples
lack measurements entirely. This distinction drives the pipeline's
automatic method selection.

The present analysis uses the non-imputed data pathway, selected on the
basis of the sensitivity analysis conducted in the preceding analysis
step. Because the cycloess-normalized data retains its natural missingness
pattern, the pipeline selects limma as the appropriate engine. This
choice preserves the biological reality that low-abundance proteins in
certain conditions are genuinely below the detection limit, rather than
filling in synthetic values that could bias fold-change estimates.

The limma pipeline applied here incorporates three layers of robustness
beyond the standard empirical Bayes framework: sample-level quality
weighting via arrayWeights, within-subject correlation modeling via
duplicateCorrelation, and outlier-robust hyperparameter estimation via
eBayes(robust = TRUE). These three mechanisms operate at different
levels---sample, design, and protein---and together produce a model that
is well-suited to the challenges of this dataset.

This document presents the results of the differential expression
analysis. All computations were performed by the companion R script
(`CvH_DEP.R`); this narrative reads the output files and presents them
with scientific context and interpretation.

The sections that follow describe the biological questions encoded as
contrasts, the statistical model and its justification, the results for
each contrast, and the multi-criteria significance framework used to
interpret findings from both well-powered and underpowered comparisons.


# 2 --- Study Contrasts: Biological Questions

The study examines skeletal muscle proteome remodeling in cancer survivors
undergoing resistance training, with a secondary question about whether
creatine supplementation modifies that response. Cancer survivors (CR)
who had completed primary treatment were randomized to receive either
creatine (CRE) or placebo (PLA) during a 12-week progressive resistance
training program. Vastus lateralis biopsies were collected at baseline
(T1) and post-training (T2). A separate group of healthy age-matched
controls (PPS) provided a single baseline biopsy for comparison.

Three contrasts capture the study's central hypotheses. Each is defined
as a linear combination of group means from the five-level factor
(CRE_T1, CRE_T2, PLA_T1, PLA_T2, H_T1) used in the design matrix.

**Training_CR** = (CRE_T2 + PLA_T2)/2 -- (CRE_T1 + PLA_T1)/2. This
contrast asks whether 12 weeks of resistance training changes the
skeletal muscle proteome in cancer survivors. By averaging across
supplement groups, it isolates the main effect of training, pooling both
CRE and PLA arms to maximize statistical power. Any protein whose
abundance shifts from T1 to T2---regardless of whether the subject
received creatine or placebo---will be detected by this contrast.

**Cancer_vs_Healthy** = (CRE_T1 + PLA_T1)/2 -- H_T1. This contrast
compares the baseline muscle proteome of cancer survivors with that of
healthy controls. Only baseline timepoints are used, ensuring that the
comparison reflects the pre-training state. Because this contrast
captures fundamental biological differences between cancer survivors and
healthy individuals---differences shaped by the disease itself, prior
treatments (chemotherapy, radiation), and associated deconditioning---it
is expected to yield the strongest signal among the three comparisons.

**Supplement_Interaction** = (CRE_T2 -- CRE_T1) -- (PLA_T2 -- PLA_T1).
This interaction contrast tests whether creatine supplementation
modifies the training response differently than placebo. It compares
the within-subject change in the CRE arm against the within-subject
change in the PLA arm. A significant result would indicate that the
proteome-level training adaptation depends on supplement assignment. This
is the most demanding contrast statistically, as it requires detecting
a difference of differences with approximately 8 subjects per supplement
group.

These three contrasts were chosen to address the study's primary aims
comprehensively. The Training_CR contrast addresses the rehabilitation
question (does exercise help?), the Cancer_vs_Healthy contrast quantifies
the deficit (how different are cancer survivors?), and the
Supplement_Interaction contrast evaluates the nutritional intervention
(does creatine add benefit?). Together they span the study's biological
scope from descriptive comparison to mechanistic interaction.

The contrast definitions are encoded in the design matrix and passed to
limma's contrasts.fit function, which re-parameterizes the fitted
coefficients to test each linear combination. Importantly, all three
contrasts are tested from the same model fit, ensuring consistent
variance estimation across comparisons.


# 3 --- Statistical Model: Three Layers of Robustness

The differential expression model used here extends the standard limma
pipeline with three complementary robustness mechanisms. Each addresses a
distinct source of heterogeneity: arrayWeights handles sample-level
quality variation, duplicateCorrelation handles within-subject
correlation from the paired design, and eBayes(robust = TRUE) handles
protein-level variance outliers. Understanding why each is necessary
requires considering the specific challenges of this dataset.

## 3.1 --- arrayWeights: Sample-Level Quality Weighting

Not all samples contribute equally to the analysis. Variation in biopsy
quality, protein extraction efficiency, and LC-MS performance means that
some samples are inherently noisier than others. The traditional approach
is to identify and remove outlier samples, but this discards potentially
informative data and requires arbitrary threshold decisions.

arrayWeights (Ritchie et al., 2006) provides a principled alternative.
It estimates a quality weight for each sample by fitting a heterogeneous
error model: samples with higher residual variance receive lower weights,
while cleaner samples receive higher weights. This weighting is
incorporated into all subsequent model fitting, so noisy samples
contribute less to coefficient estimates without being removed entirely.
The approach is analogous to weighted least squares in classical
regression, but the weights are estimated from the data rather than
specified a priori.

Liu et al. (2015) demonstrated that sample-level weighting substantially
improves power in RNA-seq analyses, and the same principle applies to
proteomics data where sample quality variation is often more pronounced
due to the complexity of the LC-MS workflow. In the present dataset,
array weights ranged from well below 1.0 (downweighted samples) to above
2.0 (upweighted samples), confirming meaningful quality heterogeneity
across the cohort.

```{r array-weights}
#| label: array-weights
#| fig-cap: "Array weights per sample. Values below 1 indicate downweighted (noisier) samples; values above 1 indicate upweighted (cleaner) samples."

aw <- read_csv(file.path(data_dir, "array_weights.csv"), show_col_types = FALSE)

cat(sprintf("Array weights: min = %.3f, median = %.3f, max = %.3f\n",
            min(aw$array_weight), median(aw$array_weight), max(aw$array_weight)))

ggplot(aw, aes(x = reorder(sample_id, array_weight), y = array_weight)) +
  geom_point(size = 2.5, color = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  coord_flip() +
  labs(x = NULL, y = "Array Weight",
       title = "Sample-Level Quality Weights (arrayWeights)") +
  theme_minimal(base_size = 11)
```

```{r aw-summary}
#| label: aw-summary

aw %>%
  summarise(
    n       = n(),
    min     = min(array_weight),
    q25     = quantile(array_weight, 0.25),
    median  = median(array_weight),
    q75     = quantile(array_weight, 0.75),
    max     = max(array_weight),
    mean    = mean(array_weight),
    sd      = sd(array_weight)
  ) %>%
  kable(digits = 3, caption = "Distribution of array weights across all samples")
```

## 3.2 --- duplicateCorrelation: Paired-Design Modeling

Cancer survivor participants contributed two biopsies each (T1 and T2),
creating a paired design. Measurements from the same individual are
correlated---if a subject has generally higher protein abundances at T1,
they tend to have generally higher abundances at T2 as well. Ignoring
this correlation inflates the residual variance and reduces power,
because the model attributes within-subject consistency to random error
rather than to the paired structure.

The duplicateCorrelation function estimates a consensus within-subject
correlation across all proteins. This single parameter captures the
average degree to which repeated measurements on the same subject
co-vary. The estimated correlation is then incorporated into the linear
model via generalized least squares, properly accounting for the
non-independence of paired observations.

This is particularly important for the Training_CR and
Supplement_Interaction contrasts, both of which compare T1 and T2
measurements within the same subjects. Without duplicateCorrelation, the
effective sample size for these within-subject contrasts would be
underestimated, and the test would be conservative. The
Cancer_vs_Healthy contrast, which compares independent groups, benefits
less directly but still gains from the improved overall variance
estimation.

The consensus correlation estimated by the pipeline reflects the
proportion of total variance attributable to between-subject differences
versus within-subject (residual) variation. A high correlation indicates
that subjects differ substantially from each other but are internally
consistent across timepoints, which is favorable for paired analyses.

## 3.3 --- eBayes(robust = TRUE): Outlier-Robust Hyperparameters

Standard empirical Bayes moderation (Smyth, 2004) assumes that all
proteins share a common prior variance distribution. The moderated
t-statistic borrows strength across proteins by shrinking each protein's
variance toward this shared prior. However, a subset of proteins with
unusually high or low variance---hypervariable proteins---can distort the
prior estimate, pulling it away from the majority and degrading the
shrinkage for well-behaved proteins.

The robust option in eBayes (Phipson et al., 2016) addresses this by
using a robust estimator for the hyperparameters of the prior
distribution. Specifically, it downweights proteins whose observed
variance is far from the fitted prior, preventing a few outliers from
corrupting the shrinkage applied to the majority. This is especially
relevant in proteomics, where blood contamination proteins (hemoglobins,
serum albumin) or highly variable structural proteins can exhibit
variances orders of magnitude larger than typical proteins.

The robust approach does not remove hypervariable proteins from the
analysis; rather, it limits their influence on the shared prior while
still testing them individually. This is preferable to manual exclusion
because it is data-driven, does not require arbitrary filtering
decisions, and preserves the full protein list for downstream analyses.

## 3.4 --- How the Three Methods Complement Each Other

The three robustness mechanisms operate at different levels of the
analysis and are therefore complementary rather than redundant.
arrayWeights operates at the sample level, adjusting for the fact that
some samples are noisier than others. duplicateCorrelation operates at
the design level, capturing the paired structure among observations from
the same subject. eBayes(robust = TRUE) operates at the protein level,
preventing hypervariable proteins from distorting the empirical Bayes
shrinkage applied to all proteins.

Together, these three layers produce a model that is resilient to the
common sources of heterogeneity in clinical proteomics: uneven sample
quality, non-independent observations from repeated measures, and
extreme variance outliers among the tested proteins. This is not a
conservative approach---it is a calibrated one. By properly modeling
each source of variation, the effective degrees of freedom and variance
estimates are more accurate, which translates to both better type I
error control and improved power to detect genuine biological
differences.


# 4 --- Results: Training_CR

The Training_CR contrast tests whether 12 weeks of progressive
resistance training alters the skeletal muscle proteome in cancer
survivors. Because this contrast averages across supplement groups, it
pools all CR participants and tests the within-subject change from T1
(baseline) to T2 (post-training). The paired design---each subject
serving as their own control---removes between-subject variability and
focuses on within-subject adaptation.

```{r training-cr-table}
#| label: training-cr-table

tcr <- read_csv(file.path(data_dir, "per_contrast_results", "Training_CR.csv"),
                show_col_types = FALSE)

tcr %>%
  arrange(adj.P.Val) %>%
  head(20) %>%
  select(gene, protein, logFC, P.Value, adj.P.Val, pi_score) %>%
  mutate(across(c(logFC, pi_score), ~round(., 3)),
         P.Value   = formatC(P.Value, format = "e", digits = 2),
         adj.P.Val = formatC(adj.P.Val, format = "e", digits = 2)) %>%
  kable(caption = "Training_CR: Top 20 proteins ranked by adjusted p-value")
```

The number of differentially expressed proteins at FDR < 0.05 reflects
the modest statistical power available for within-subject contrasts when
the sample size per group is relatively small. Resistance training
induces well-characterized adaptations in skeletal muscle---including
shifts in contractile protein isoforms, mitochondrial biogenesis markers,
and extracellular matrix remodeling---but detecting these at the
proteome level requires sufficient replication to distinguish biological
signal from measurement noise.

```{r training-cr-hist}
#| label: training-cr-hist
#| fig-cap: "P-value distribution for the Training_CR contrast. A spike near zero indicates genuine signal; a uniform distribution suggests no systematic effect."

ggplot(tcr, aes(x = P.Value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", boundary = 0) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", alpha = 0.6) +
  labs(x = "Nominal P-value", y = "Count",
       title = "Training_CR --- P-value Distribution") +
  theme_minimal(base_size = 12)
```

The p-value histogram provides a diagnostic beyond the simple DEP count.
A well-calibrated test under a mixture of null and alternative hypotheses
produces a histogram with a uniform component (the null proteins) and an
enrichment near zero (the true positives). The height and sharpness of
the near-zero peak indicates the strength and prevalence of the training
effect across the proteome.

Proteins that shift with training in cancer survivors may include
contractile apparatus components (myosin heavy chains, troponins),
metabolic enzymes involved in oxidative phosphorylation and glycolysis,
and structural proteins of the extracellular matrix. The direction of
change is informative: upregulation of mitochondrial and contractile
proteins would suggest successful adaptation, while persistent
downregulation might indicate incomplete recovery from cancer-related
deconditioning.

The volcano plot below visualizes the relationship between effect size (log2 fold change, x-axis) and statistical significance (-log10 p-value, y-axis) for the Training_CR contrast. Each dot represents one protein. Proteins in the upper-left and upper-right corners show both large fold changes and high statistical significance. Red dots are significantly upregulated after training; blue dots are significantly downregulated. Gene names are shown for proteins reaching FDR < 0.05.

```{r training-cr-volcano}
#| label: training-cr-volcano
#| fig-cap: "Volcano plot for the Training_CR contrast. Red = significantly upregulated (FDR < 0.05, logFC > 0); Blue = significantly downregulated (FDR < 0.05, logFC < 0); Gray = not significant."

tcr_volc <- tcr %>%
  mutate(sig = case_when(
    adj.P.Val < 0.05 & logFC > 0 ~ "Up",
    adj.P.Val < 0.05 & logFC < 0 ~ "Down",
    TRUE ~ "NS"
  ))

sig_counts <- tcr_volc %>% count(sig)
up_n   <- sig_counts %>% filter(sig == "Up") %>% pull(n) %>% { if (length(.) == 0) 0L else . }
down_n <- sig_counts %>% filter(sig == "Down") %>% pull(n) %>% { if (length(.) == 0) 0L else . }
ns_n   <- sig_counts %>% filter(sig == "NS") %>% pull(n) %>% { if (length(.) == 0) 0L else . }

volc_colors <- c("Up" = "#B2182B", "Down" = "#2166AC", "NS" = "gray70")
volc_labels <- c(
  "Up"   = paste0("Up (", up_n, ")"),
  "Down" = paste0("Down (", down_n, ")"),
  "NS"   = paste0("NS (", ns_n, ")")
)

ggplot(tcr_volc, aes(x = logFC, y = -log10(P.Value), color = sig)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_text_repel(
    data = tcr_volc %>% filter(adj.P.Val < 0.05),
    aes(label = gene),
    size = 2.5, max.overlaps = 20, show.legend = FALSE
  ) +
  scale_color_manual(values = volc_colors, labels = volc_labels) +
  labs(x = "log2 Fold Change", y = "-log10(P-value)",
       title = "Training_CR --- Volcano Plot",
       color = "Significance") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Even when few proteins cross the FDR < 0.05 threshold, the directional
trends and top-ranked proteins are informative. The pi-score (described
in Section 7) provides an alternative ranking that integrates fold change
and significance, which can be particularly useful for generating
hypotheses when formal significance is limited by sample size.


# 5 --- Results: Cancer_vs_Healthy

The Cancer_vs_Healthy contrast compares the baseline skeletal muscle
proteome of cancer survivors with that of healthy age-matched controls.
This comparison captures the cumulative effects of cancer, prior
treatment, and associated physical deconditioning on the muscle
proteome. Because it contrasts two independent groups (CR at T1 versus
PPS at T1), it does not benefit from the paired design and relies
instead on the between-group sample sizes.

```{r cvh-table}
#| label: cvh-table

cvh <- read_csv(file.path(data_dir, "per_contrast_results", "Cancer_vs_Healthy.csv"),
                show_col_types = FALSE)

cvh %>%
  arrange(adj.P.Val) %>%
  head(20) %>%
  select(gene, protein, logFC, P.Value, adj.P.Val, pi_score) %>%
  mutate(across(c(logFC, pi_score), ~round(., 3)),
         P.Value   = formatC(P.Value, format = "e", digits = 2),
         adj.P.Val = formatC(adj.P.Val, format = "e", digits = 2)) %>%
  kable(caption = "Cancer_vs_Healthy: Top 20 proteins ranked by adjusted p-value")
```

This contrast is expected to yield the largest number of differentially
expressed proteins among the three comparisons. The biological distance
between cancer survivors and healthy controls encompasses not only the
direct effects of malignancy but also the downstream consequences of
treatment (chemotherapy-induced muscle wasting, corticosteroid exposure,
prolonged inactivity) and the systemic inflammatory state that persists
in many survivors. These are profound perturbations that leave measurable
imprints on the muscle proteome.

```{r cvh-hist}
#| label: cvh-hist
#| fig-cap: "P-value distribution for the Cancer_vs_Healthy contrast."

ggplot(cvh, aes(x = P.Value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", boundary = 0) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", alpha = 0.6) +
  labs(x = "Nominal P-value", y = "Count",
       title = "Cancer_vs_Healthy --- P-value Distribution") +
  theme_minimal(base_size = 12)
```

The p-value histogram for this contrast should show a pronounced spike
near zero, reflecting the abundance of truly differentially expressed
proteins. A strong leftward skew confirms that the biological signal
overwhelms the null distribution and that the multiple testing correction
retains a substantial number of discoveries at FDR < 0.05.

Biologically, proteins elevated in cancer survivors relative to healthy
controls may include acute-phase reactants, stress-response proteins,
and markers of proteolytic pathway activation (ubiquitin-proteasome,
autophagy). Proteins depleted in cancer survivors may include
mitochondrial enzymes, contractile machinery components, and metabolic
enzymes associated with healthy muscle function. The pattern of
up-and-down regulation can reveal which pathways are activated versus
suppressed in the cancer survivor muscle, providing a molecular
fingerprint of cancer-related muscle dysfunction.

The volcano plot below shows the magnitude and significance of proteomic differences between cancer survivors and healthy controls at baseline. Proteins in the upper corners represent the strongest candidates for cancer-associated muscle dysfunction. Red dots indicate proteins elevated in cancer survivors relative to healthy controls; blue dots indicate proteins depleted in cancer survivors.

```{r cvh-volcano}
#| label: cvh-volcano
#| fig-cap: "Volcano plot for the Cancer_vs_Healthy contrast. Red = significantly upregulated in cancer survivors (FDR < 0.05, logFC > 0); Blue = significantly downregulated (FDR < 0.05, logFC < 0); Gray = not significant."

cvh_volc <- cvh %>%
  mutate(sig = case_when(
    adj.P.Val < 0.05 & logFC > 0 ~ "Up",
    adj.P.Val < 0.05 & logFC < 0 ~ "Down",
    TRUE ~ "NS"
  ))

sig_counts <- cvh_volc %>% count(sig)
up_n   <- sig_counts %>% filter(sig == "Up") %>% pull(n) %>% { if (length(.) == 0) 0L else . }
down_n <- sig_counts %>% filter(sig == "Down") %>% pull(n) %>% { if (length(.) == 0) 0L else . }
ns_n   <- sig_counts %>% filter(sig == "NS") %>% pull(n) %>% { if (length(.) == 0) 0L else . }

volc_colors <- c("Up" = "#B2182B", "Down" = "#2166AC", "NS" = "gray70")
volc_labels <- c(
  "Up"   = paste0("Up (", up_n, ")"),
  "Down" = paste0("Down (", down_n, ")"),
  "NS"   = paste0("NS (", ns_n, ")")
)

ggplot(cvh_volc, aes(x = logFC, y = -log10(P.Value), color = sig)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_text_repel(
    data = cvh_volc %>% filter(adj.P.Val < 0.05),
    aes(label = gene),
    size = 2.5, max.overlaps = 20, show.legend = FALSE
  ) +
  scale_color_manual(values = volc_colors, labels = volc_labels) +
  labs(x = "log2 Fold Change", y = "-log10(P-value)",
       title = "Cancer_vs_Healthy --- Volcano Plot",
       color = "Significance") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Interpreting this contrast requires caution about confounding.
Differences between CR and PPS groups may reflect not only cancer and
treatment effects but also unmeasured lifestyle, dietary, or genetic
differences between the groups. The study design mitigates this through
age matching, but residual confounding cannot be excluded in an
observational comparison. The results should therefore be interpreted as
associative rather than causal.


# 6 --- Results: Supplement_Interaction

The Supplement_Interaction contrast tests whether creatine
supplementation modifies the training-induced proteome response
differently than placebo. As an interaction contrast, it compares the
within-subject change (T2 minus T1) in the CRE arm against the
within-subject change in the PLA arm. A significant result would
indicate that the proteome-level training adaptation depends on
supplement group assignment.

```{r supp-table}
#| label: supp-table

supp <- read_csv(file.path(data_dir, "per_contrast_results", "Supplement_Interaction.csv"),
                 show_col_types = FALSE)

supp %>%
  arrange(adj.P.Val) %>%
  head(20) %>%
  select(gene, protein, logFC, P.Value, adj.P.Val, pi_score) %>%
  mutate(across(c(logFC, pi_score), ~round(., 3)),
         P.Value   = formatC(P.Value, format = "e", digits = 2),
         adj.P.Val = formatC(adj.P.Val, format = "e", digits = 2)) %>%
  kable(caption = "Supplement_Interaction: Top 20 proteins ranked by adjusted p-value")
```

This contrast is the most statistically demanding of the three. Testing
an interaction requires detecting a difference of differences, which
effectively halves the signal relative to a main-effect contrast. With
approximately 8 subjects per supplement group, power to detect moderate
interaction effects is limited. The absence of significant findings at
FDR < 0.05 does not necessarily mean that creatine has no effect on the
training response---it may simply mean that the study was underpowered
to detect such effects at the proteome level.

```{r supp-hist}
#| label: supp-hist
#| fig-cap: "P-value distribution for the Supplement_Interaction contrast. A uniform distribution indicates no systematic signal."

ggplot(supp, aes(x = P.Value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", boundary = 0) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red", alpha = 0.6) +
  labs(x = "Nominal P-value", y = "Count",
       title = "Supplement_Interaction --- P-value Distribution") +
  theme_minimal(base_size = 12)
```

The p-value histogram is the most informative diagnostic for
underpowered contrasts. A uniform distribution across the 0-to-1 range
confirms that there is no systematic departure from the null hypothesis:
the observed p-values are consistent with what would be expected if all
proteins were truly null. This pattern---a flat histogram---is a
hallmark of a well-calibrated test applied to data with no signal,
rather than an indication of analytical failure.

Distinguishing a true biological null from an underpowered analysis
requires external information. Creatine supplementation during
resistance training has shown mixed results in the broader literature,
with some studies reporting benefits in lean mass accretion and strength
gains but fewer demonstrating proteome-level effects. The present sample
size of approximately 8 per group is near the lower bound of what is
detectable for proteomics interaction effects, so the null result should
be interpreted conservatively.

The volcano plot below displays the interaction effect --- whether creatine supplementation modifies the training-induced proteome response differently than placebo. With the limited sample size for this interaction contrast, few or no proteins may reach formal significance (FDR < 0.05), but the overall shape of the volcano reveals whether there are any trends toward differential supplement effects.

```{r supp-volcano}
#| label: supp-volcano
#| fig-cap: "Volcano plot for the Supplement_Interaction contrast. Red = significantly upregulated (FDR < 0.05, logFC > 0); Blue = significantly downregulated (FDR < 0.05, logFC < 0); Gray = not significant."

supp_volc <- supp %>%
  mutate(sig = case_when(
    adj.P.Val < 0.05 & logFC > 0 ~ "Up",
    adj.P.Val < 0.05 & logFC < 0 ~ "Down",
    TRUE ~ "NS"
  ))

sig_counts <- supp_volc %>% count(sig)
up_n   <- sig_counts %>% filter(sig == "Up") %>% pull(n) %>% { if (length(.) == 0) 0L else . }
down_n <- sig_counts %>% filter(sig == "Down") %>% pull(n) %>% { if (length(.) == 0) 0L else . }
ns_n   <- sig_counts %>% filter(sig == "NS") %>% pull(n) %>% { if (length(.) == 0) 0L else . }

volc_colors <- c("Up" = "#B2182B", "Down" = "#2166AC", "NS" = "gray70")
volc_labels <- c(
  "Up"   = paste0("Up (", up_n, ")"),
  "Down" = paste0("Down (", down_n, ")"),
  "NS"   = paste0("NS (", ns_n, ")")
)

ggplot(supp_volc, aes(x = logFC, y = -log10(P.Value), color = sig)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_text_repel(
    data = supp_volc %>% filter(adj.P.Val < 0.05),
    aes(label = gene),
    size = 2.5, max.overlaps = 20, show.legend = FALSE
  ) +
  scale_color_manual(values = volc_colors, labels = volc_labels) +
  labs(x = "log2 Fold Change", y = "-log10(P-value)",
       title = "Supplement_Interaction --- Volcano Plot",
       color = "Significance") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

For underpowered contrasts where formal FDR control yields no
discoveries, the pi-score provides an alternative ranking that
integrates both effect size and statistical evidence. Proteins with high
pi-scores---those with large fold changes and low p-values, even if
neither alone reaches a strict threshold---represent the most promising
candidates for follow-up in larger studies. The pi-score approach is
discussed in detail in Section 7.


# 7 --- Multi-Criteria Significance

Proteomics experiments with limited sample sizes benefit from a
multi-criteria significance framework that goes beyond a single FDR
threshold. Different thresholds serve different scientific purposes, and
the appropriate choice depends on the goal of the analysis: publication-
ready claims require strict control, while hypothesis generation for
follow-up studies can tolerate more liberal criteria.

```{r da-summary}
#| label: da-summary

da <- read_csv(file.path(data_dir, "DA_summary.csv"), show_col_types = FALSE)

da %>%
  kable(caption = "Differential abundance summary across contrasts and significance criteria")
```

**FDR < 0.05** is the gold standard for publication-ready differential
expression claims. The Benjamini-Hochberg procedure controls the expected
proportion of false discoveries among all proteins declared significant.
At this threshold, no more than 5% of the reported hits are expected to
be false positives on average. This is the appropriate criterion for
making confident biological claims and for prioritizing proteins for
expensive validation experiments (e.g., Western blot, targeted MS).

**FDR < 0.10** is an exploratory threshold acceptable for hypothesis
generation. It relaxes the false discovery control slightly, allowing
more proteins through at the cost of a modestly higher false positive
rate. This threshold is useful for pathway-level analyses where the
goal is to identify enriched biological processes rather than to make
claims about individual proteins. The logic is that pathway enrichment
analysis aggregates information across many proteins, so individual
false positives are less damaging when the analytical unit shifts from
proteins to pathways.

**Nominal p < 0.01** bypasses multiple testing correction entirely and
is appropriate only when combined with effect-size thresholds or when
the goal is exploratory ranking. For underpowered contrasts such as the
Supplement_Interaction, applying strict FDR control may yield zero
discoveries even when biologically meaningful trends exist. The nominal
threshold, combined with fold-change filtering, can identify proteins
worthy of follow-up while clearly acknowledging the increased risk of
false positives. Results reported at this threshold should be framed as
hypothesis-generating rather than confirmatory.

**Pi-score** (Xiao et al., 2014) provides an unbiased ranking that
integrates statistical significance and biological significance (effect
size) into a single metric. The pi-score is defined as
$-\log_{10}(p) \times |\text{logFC}|$, which gives high scores to
proteins that are both statistically significant and biologically
meaningful. This is particularly valuable for underpowered contrasts
where neither p-value nor fold change alone provides a reliable ranking:
a protein with a modest p-value and a large fold change may be more
biologically interesting than one with a slightly lower p-value but
negligible effect size.

The pi-score ranks are converted to empirical p-values by ranking the
scores and dividing by the total number of proteins tested. The top 5%
by pi-score (pi_pval < 0.05) represent proteins with the strongest
combined evidence of statistical and biological significance. This
approach is especially useful for the Supplement_Interaction contrast,
where no proteins reach FDR < 0.05 but the top pi-score candidates
may still point to biologically coherent pathways.

The summary table above reports counts under each criterion for all
three contrasts, stratified by direction (up, down, nonsignificant).
The Cancer_vs_Healthy contrast dominates in terms of discovery count
at every threshold, consistent with its expected biological strength.
The Training_CR contrast yields moderate discoveries, and the
Supplement_Interaction contrast yields few or none under stringent
criteria. This pattern is consistent with the expected effect sizes
and the available statistical power for each comparison.

Across all criteria, it is essential to interpret the counts in the
context of the total number of proteins tested. A discovery count of
14 out of 1,829 tested proteins at FDR < 0.05 carries different
biological weight than 14 out of 100. The per-contrast results tables
(Sections 4--6) provide the individual protein identities behind these
counts, enabling biological interpretation beyond summary statistics.


# 8 --- Conclusion

The differential expression analysis of the CvH proteomics dataset
reveals a graded pattern of signal strength across the three study
contrasts. The Cancer_vs_Healthy comparison yields the strongest and
most abundant differential expression, consistent with the profound
biological distance between cancer survivors and healthy controls. The
Training_CR contrast identifies a more modest set of proteins that
respond to 12 weeks of resistance training, reflecting the within-subject
paired design and the averaging across supplement groups. The
Supplement_Interaction contrast yields no discoveries at FDR < 0.05,
consistent with either a true absence of creatine-specific proteome
effects or insufficient power to detect them with the available sample
size.

The three-layer robustness approach---arrayWeights, duplicateCorrelation,
and robust eBayes---provides a model that is well-calibrated for the
specific challenges of this clinical proteomics dataset. Sample quality
weighting avoids the information loss of outlier removal, paired-design
modeling captures the within-subject correlation structure, and robust
hyperparameter estimation protects against the influence of hypervariable
proteins. Together, these mechanisms maximize the extractable signal
while maintaining rigorous type I error control.

The multi-criteria significance framework accommodates the different
power profiles of the three contrasts. For Cancer_vs_Healthy, strict FDR
control is appropriate and productive. For Training_CR, a combination of
FDR < 0.05 and exploratory FDR < 0.10 thresholds provides a tiered view
of the training response. For Supplement_Interaction, pi-score ranking
and nominal p-value thresholds are the most informative tools for
identifying candidate proteins worthy of further investigation.

These results establish the foundation for downstream analyses, including
pathway enrichment, network analysis, and integration with clinical
outcome data. The per-contrast results files and the combined results
table provide the inputs needed for these subsequent analyses.


# References

1. Ritchie ME, Phipson B, Wu D, et al. (2015). limma powers differential
   expression analyses for RNA-sequencing and microarray studies. *Nucleic
   Acids Research* 43(7):e47.

2. Phipson B, Lee S, Oshlack A, et al. (2016). Robust hyperparameter
   estimation protects against hypervariable genes and improves power to
   detect differential expression. *Annals of Applied Statistics*
   10(2):946-963.

3. Zhu Y, Orre LM, Johansson HJ, et al. (2020). DEqMS: a method for
   accurate variance estimation in differential protein expression
   analysis. *Molecular & Cellular Proteomics* 19(6):1047-1057.

4. Ritchie ME, Diyagama D, Neilson J, et al. (2006). Empirical array
   quality weights in the analysis of microarray data. *BMC
   Bioinformatics* 7:261.

5. Smyth GK (2004). Linear models and empirical Bayes methods for
   assessing differential expression in microarray experiments.
   *Statistical Applications in Genetics and Molecular Biology*
   3(1):Article 3.

6. Xiao Y, Hsiao TH, Suez U, et al. (2014). A novel significance score
   for gene selection and ranking. *Bioinformatics* 30(6):801-807.

7. Liu R, Holik AZ, Su S, et al. (2015). Why weight? Modelling sample
   and observational level variability improves power in RNA-seq
   analyses. *Nucleic Acids Research* 43(15):e97.

8. Valikangas T, Suomi T, Elo LL (2018). A systematic evaluation of
   normalization methods in quantitative label-free proteomics. *Briefings
   in Bioinformatics* 19(1):1-11.


# Session Info

```{r session-info}
#| label: session-info

sessionInfo()
```
