---
title: "CvH 02 — Normalization & Imputation"
subtitle: "Cyclic-loess normalization, and when to fill missing values"
author: "A_CvH_2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    embed-resources: true
    df-print: paged
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
pacman::p_load(here)
knitr::opts_knit$set(root.dir = here::here())
```

This document reproduces stage 02 of the CvH pipeline and shows the result of
each step. It runs the same logic as `02_Normalization/a_script/01_run_normalization.R`
(plus the three imputation arms in `02_Normalization/imputation/a_script/`), in
memory, writing no files. Normalization makes samples comparable; imputation is
applied only for the methods that need a complete matrix.

# Load the filtered `DAList` {#sec-load}

Stage 01 hands off a filtered, un-normalized object that still carries its
missing values:

```{r}
#| label: load
pacman::p_load(proteoDA)
dal <- readRDS(here("01_Filtering", "c_data", "DAList_filtered.rds"))
miss_before <- round(100 * mean(is.na(dal$data)), 1)
dim(dal$data)
miss_before # ~17% missing — not yet imputed
dal$tags # provenance so far: the group filter
```

# Normalize with cyclic loess {#sec-normalize}

Samples load onto the instrument slightly differently, so whole samples sit
higher or lower for technical reasons. This shows in each sample's median
log-intensity. `normalize_data(norm_method = "cycloess")` evens it out with a
single native call:

```{r}
#| label: normalize
before <- apply(log2(as.matrix(dal$data)), 2, median, na.rm = TRUE)
dal <- normalize_data(dal, norm_method = "cycloess")
after <- apply(as.matrix(dal$data), 2, median, na.rm = TRUE)

rbind(before = summary(before), after = summary(after)) # spread collapses to a tight band
dal$tags # new 'normalized' + 'norm_method' tags
```

Normalization **rescales** but does not fill anything in — the missingness is
unchanged:

```{r}
#| label: missing-unchanged
c(
  before_pct = miss_before,
  after_pct = round(100 * mean(is.na(dal$data)), 1)
)
```

The script also writes proteoDA's before/after audit PDFs to `b_reports/`. Those
calls (shown here, run by the script) make the normalization choice reviewable:

```{r}
#| label: qc-reports
#| eval: false
write_norm_report(dal,
  grouping_column = "group_time", output_dir = report_dir,
  filename = "norm_comparison.pdf"
) # compares candidate methods (before normalize)
write_qc_report(dal,
  color_column = "group_time", output_dir = report_dir,
  filename = "qc_normalized.pdf"
) # inspects the chosen result
```

This normalized, **non-imputed** object is the primary hand-off to stage 03 —
`limma` handles missing values per protein, so imputing before the test would
inflate false positives.

# Imputation for methods that need a complete matrix {#sec-impute}

PCA, clustering, WGCNA, and the reversal rotation tests need a matrix with no
gaps. Two independent imputation methods are kept so the choice can be checked.
Both operate on the normalized matrix and fill every missing value.

## imp4p (canonical) {#sec-imp4p}

`imp4p` is mechanism-aware: it estimates the per-protein mix of values missing at
random vs missing because the protein is too low to detect, then imputes each
accordingly. It is stochastic (multiple imputation), so the seed is fixed:

```{r}
#| label: imp4p
#| cache: true
pacman::p_load(imp4p)
set.seed(42)
mat <- as.matrix(dal$data)
cond <- factor(dal$metadata$group_time[match(colnames(mat), dal$metadata$Col_ID)])
imp4p_mat <- as.matrix(impute.mi(
  tab = mat, conditions = cond,
  methodMCAR = "mle", methodMNAR = "igcda", progress.bar = FALSE
))
dimnames(imp4p_mat) <- dimnames(mat)

c(
  dim_proteins = nrow(imp4p_mat), dim_samples = ncol(imp4p_mat),
  na_remaining = sum(is.na(imp4p_mat))
) # 0 remaining — complete matrix
```

## MsCoreUtils hybrid (second opinion) {#sec-mscoreutils}

The second method classifies each protein missing-at-random (MAR) vs not via
`imputeLCMD::model.Selector()`, then routes MAR features to kNN and MNAR features
to a left-censored QRILC draw:

```{r}
#| label: mscoreutils
pacman::p_load(MsCoreUtils, imputeLCMD)
set.seed(42)
randna <- as.logical(model.Selector(mat)[[1]]) # TRUE = MAR feature
ms_mat <- impute_matrix(mat, method = "mixed", randna = randna, mar = "knn", mnar = "QRILC")

c(
  MAR_features = sum(randna), MNAR_features = sum(!randna),
  na_remaining = sum(is.na(ms_mat))
)
```

## missForest (random-forest comparison) {#sec-missforest}

The third arm is mechanism-agnostic: `missForest` predicts each missing value from
the multivariate structure of the other samples, with no left-censoring model. It
tracks the non-imputed effect sizes closely and is kept alongside the other two as
a comparison:

```{r}
#| label: missforest
pacman::p_load(missForest)
set.seed(42)
mf_mat <- missForest(mat, maxiter = 10, ntree = 100, verbose = FALSE)$ximp
dimnames(mf_mat) <- dimnames(mat)

c(
  dim_proteins = nrow(mf_mat), dim_samples = ncol(mf_mat),
  na_remaining = sum(is.na(mf_mat))
)
```

The methods only differ where a value was actually missing — a modest slice
of the matrix:

```{r}
#| label: compare
was_missing <- is.na(mat)
c(
  entries_imputed = sum(was_missing),
  pct_of_matrix = round(100 * mean(was_missing), 1)
)
```

# Where stage 02 stands {#sec-recap}

The normalized matrix is saved to `02_Normalization/c_data/`; the three imputation
arms write to `02_Normalization/imputation/c_data/`:

| File | Location | Missing values? | Used by |
|---|---|---|---|
| `DAList_normalized.rds` | `c_data/` | kept | the DE test (stage 03) — **primary** |
| `DAList_imputed_imp4p.rds` | `imputation/c_data/` | filled | figures, rotation tests (canonical) |
| `DAList_imputed_mscoreutils.rds` | `imputation/c_data/` | filled | the robustness comparison |
| `DAList_imputed_missforest.rds` | `imputation/c_data/` | filled | the robustness comparison |

Normalization levels the samples but never invents data; imputation is applied
only for methods that cannot run on an incomplete matrix, and never before the DE
test. Stage 03 fits the model on the non-imputed matrix.
